Weak interspecific interactions in a sagebrush steppe: evidence from observations, models, and experiments

### 2017-09-09 removals/manuscript/ResponseToReviews.tex

To clarify these issues, we have DO THIS added new text to Methods??? Results??? Discussion???, including the figure showing the marginal effects of A. tripartita crowding. Should we also mention how we would do it differently next time, in Discussion???}

@@

Possibly related to this – where the removal and competition parameters correlated?

\response{Not sure what the reviewer means...perhaps correlated in the MCMC chains?}

Another thing that kept crossing my mind as I read the paper was about the treatment of “non-significant” parameters, i.e., parameters with high uncertainty. As far as I could tell throughout the paper, parameters were never dropped from models, and projections were done using the parameter estimates based on the mean of the MCMC chain values. I was struggling with what it means to use parameters with high uncertainty in projections. You explain that you want to use the deterministic versions of the models (line 186) rather than a draw from the Poisson random variable; however those deterministic values of lambda have uncertainty. In your simulations you could incorporate that uncertainty by taking a random draw from the MCMC chain for each parameter (I don't know if this is standard practice). It seemed that this would somehow “equalize” parameter uncertainty (in competition, yearly intercept, removal parameters, etc.). I’m not sure if this would change the conclusions, the only thing it might do is make the variance larger in Fig. 5, and thus the P. spicata Removal model/no ARTR might not look as different from the other P. spicata models. I’m not saying you need to do this, I’m just curious if you had considered this issue at all.

\response{ This is a good point, and we appreciate being pushed to justify our decision to ignore parameter uncertainty. Our goal is to test
whether competitive release is as small as our previous model predicted. To be conservative, we want to make it as easy as possible to
model large effects of removals. (That is why we included the "maximum removal effect" simulation based on the outer limits of the removal effect
confidence intervals.) Propagating parameter uncertainty through our simulations would add noise, making it \emph{harder} to 
detect the effects of removals. More formally, our previous result showed that the interspecific competition coefficients are small in an ecologically
relevant sense: the shift in mean cover due to interspecific competition shown in Fig. 5 is (for 3 species) small compared to the natural fluctuations in cover driven by interannual environmental variability. 
If, as suggested, we made the projections with parameter uncertainty on top of random year effects, we would be conflating parameter uncertainty with environmental variability, ``tilting'' the
comparison in our favor by overstating the magnitude of natural fluctuations. One point of Fig. 5 is that for \emph{P. spicata} the small effect of competition with ARTR each year adds up (in the model) to a nontrivial long-term effect.

@@

\response{TO DO: add caveat}

Given this result for P. spicata, line 401 of the discussion is a bit surprising: “how to reconcile the
insensitivity of grasses to sagebrush removal…..” On the other hand, it does hold for two of the grasses
and some discussion of why the grass species differ in sensitivity to shrub removal would be interesting.
Regardless of which species it holds for, the authors then attempt to reconcile the insensitivity found in
this study with strong effects found in the rangeland ecology literature by noting the relatively low shrub
cover (and small size of A. tripartita) in these ungrazed areas to explain the lack of strong effect of shrub
removal on grasses (lines 401-412). I agree this is a reasonable explanation and would argue that this is
also a possible, even likely, explanation for lack of effect of the grasses on sagebrush as above. In either
case, this reason for weak interspecific interactions (few interspecific neighbors) is quite different from
the ecological explanations in the next paragraph (lines 413-426; e.g., resource partitioning, balance of
facilitation and competition) and has quite different implications from those discussed in the
subsequent paragraph (lines 427-438). If this material stays, it would be helpful to compare and
contrast these different sorts of explanations and their implications more directly.

\response{We thank the reviewer for pointing out this apparent contradiction. In fact, consistent with 
the reviewers' critique of our argument, we actually demonstrated that 
artificially increasing Artemisia cover would not lead to a much bigger competitive release. We have revised this
section to suggest that overgrazing must do more than simply increase shrub cover, but must also alter
the per capita competition coefficients, potentially by altering the balance between competition and facilitation
or changing resource availability. DO THIS}

@@

A few minor points:

Lines 50-53: Are these two sentences partly saying the same thing? If there is full compensation, the
response must depend on how much was removed, i.e., the degree of dominance.

\response{We revised this passage to make it clear that the second sentence involves partial, not full
compensation.}

@@

Line 81: should “determine” be replaced by “estimate”?

\response{We have added "estimated" later in the sentence to recognize the uncertainty in the model.}

@@

Line 87: given that you earlier defined competitive release as a population response, should
“competitive release” at the end of this sentence be replaced by “individual response to neighbors”?
The second question more directly addresses the population consequences.

\response{We went with ``underestimate the response to decreased crowding" because our recruitment model does not operate 
at the level of individuals.}

@@

Line 216: “very similar to those used in our previous studies” is a bit confusing. Why not just use the
vital rate functions calculated in the first part of the study? If they are indeed the same, say so rather
than introducing similarity to other published studies again.

\response{We do use the vital rate functions calculated in the first part of \emph{this} study, but they differ in very subtle ways 
from the models used in our previous studies. We have revised the sentence to clarify this.}

@@

Line 234-235: Several confusing points here: a) The distinction between the community reaching
equilibrium (500 time steps) and each species reaching equilibrium cover (“additional 2000 time steps”)
is not clear. b) I assume the random year effects were used for all 2500 time steps? c) were the 3
scenarios applied AFTER the community reached equilibrium but before the 2000 steps, i.e., removal
from an equilibrium community? Or from the beginning of each set of runs?

\response{ This was bad phrasing on our part. 
The community equilibrium is not a constant level of cover for each species, as it would be in a deterministic
model, but a steady-state pattern of fluctuations over time. The additional 2000 time steps are needed to compute the average cover of each species
at the community steady-state. We have re-phrased and expanded these sentences to clarify this point. b) Yes, that's right. The paragraph
has been reorganized to clarify this point. c) The scenarios were applied from the very beginning; the paragraph has been re-organized to 
clarify this point.}

Fig. 4. Is the baseline removal line missing for H. comata?

\response{It is there, but virtually indistinguishable from the removal effect line; we've added text to the legend to point this out.}

### 2017-09-09 removals/manuscript/removal_ms.tex
More revisions...

An alternative approach to studying competitive release is to fit multispecies population models that estimate species interactions using long-term, observational data, and then simulate population responses to removals. Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and weak interspecific interactions, leading to very stable coexistence and implying little competitive release \citep{adler_coexistence_2010,chu_large_2015}. In fact, at our sagebrush steppe study site  \citep{adler_coexistence_2010}, our model predicts removal of sagebrush would actually reduce the equilibrium cover of the three dominant grass species very slightly, reflecting the facilitative effect of sagebrush on grass recruitment and very weak competitive effects on growth and survival.  Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but the predicted negative effect of sagebrush removal on grass cover is surprising given the long history of sagebrush eradication to increase grass forage production in sagebrush rangelands \citep{Cook1963, Mueggler1958, Robertson1947}.
Compared to traditional field experiments, our approach has the advantage of projecting full life-cycle responses over long time periods, but its reliance on observational data sets limits the strength of our inference. We would have more confidence in our model's projections of competitive release if we could experimentally confirm the ability of our models to describe interspecific interactions.

To test whether our sagebrush community model, which is based on observational data, underestimates the strength of interspecific competition and competitive release, we conducted a removal experiment in a sagebrush steppe. We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses. The primary purpose of the experiment was not to directly measure competitive release, which is difficult to do over a short time period in a semiarid ecosystem with low and variable plant cover, but rather to validate the parameters in our model that represent key interspecific interactions and ultimately determine the magnitude of competitive release. We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs. removal treatments after accounting for differences in local species composition? Because our demographic models include the effects of neighborhood crowding, they should be able to predict how experimental removal of neighbors alters demographic performance. If the full effect of experimental removals is stronger than predicted, it means that our statistical models underestimate competitive release. Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data? If models based on observational data adequately describe facilitative and competitive interactions, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments, a result which would validate the use of models and long-term data sets to study interspecific interactions.

@@

The kernel is constructed from the fitted survival ($S$), growth ($G$), and recruitment ($R$) models, 
\begin{equation}
k_j(v,u,\vec{n})=S_j(u,\vec{n})G_j(v,u,\vec{n})+R_j(v,u,\vec{n}) 
\label{eqn:IPM} 
\end{equation}
which are very similar to those used in our previous studies of this system \citep{adler_coexistence_2010,adler_forecasting_2012}; see 
Appendix \ref{AppendixA} for a complete description. As in these previous studies, the IPM uses a mean-field approximation for
neighborhood competition that captures the essential features of spatially-explicit local interactions; 
in Appendix \ref{sec:kernelMethods} we explain how the spline competition kernels were used 
to compute neighborhood competition as a function of individual size for the IPM.

@@

 \begin{figure}[tbp]
 \centering
 \includegraphics[width=1\textwidth]{cover_projections_1step}
 \caption{Observed and predicted cover for each of the four modeled species. Solid black lines show mean observed cover averaged across control (closed symbols) and removal treatment (open symbols) quadrats. Blue lines and symbols show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats. Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients. Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}. }
 \label{fig:IBM1step}
 \end{figure}

@@

@@

Evidence for weak interspecific interactions and limited competitive release has interesting implications. If most 
interspecific interactions are weak, then single-species models may be adequate for forecasting population responses 
to environmental change \citep{adler_forecasting_2012,kleinhesselink_indirect_2015,Chu2016}. Similarly, differences among 
species in relative abundance should reflect variation in the strength of intraspecific density dependence more than 
variation in the strength of interspecific interactions \citep{yenni_strong_2012}. A recent global study showing little 
variation in the abundance of invasive grassland species in native and invaded ranges on different 
continents  \citep{firn_abundance_2011} is consistent with the idea that intraspecific interactions are the primary determinant 
of species abundances. In systems where this is true, effective management and conservation requires understanding the sources 
and species-specificity of intraspecific density dependence. A useful first step in this direction would be to test 
whether our model is also successful at predicting the effects of release from intraspecific competition

>>>>

An alternative approach to studying competitive release is to fit multispecies population models that estimate species interactions using long-term, observational data, and then simulate population responses to removals. Applying this approach to arid and semiarid plant communities, we found evidence of strong intraspecific limitation and weak interspecific interactions, leading to very stable coexistence and implying little competitive release \citep{adler_coexistence_2010,chu_large_2015}. In fact, at our sagebrush steppe study site  \citep{adler_coexistence_2010}, our model predicts removal of sagebrush would actually reduce the equilibrium cover of the three dominant grass species very slightly, reflecting the facilitative effect of sagebrush on grass recruitment and very weak competitive effects on growth and survival.  Since we only studied the most common, co-occurring species, we expected to find evidence of stable coexistence, but the predicted negative effect of sagebrush removal on grass cover is surprising given the long history of sagebrush eradication to increase grass forage production in sagebrush rangelands \citep{Cook1963, Mueggler1958, Robertson1947}.
Compared to traditional field experiments, our approach has the advantage of projecting full life-cycle responses over long time periods, but its reliance on observational data sets limits the strength of our inference. A successful experimental test could increase our confidence in the model's projections of competitive release.

\new{ The idea that results from observational studies of competition represent hypotheses which should be tested experimentally is not new (Hairston 1980), but such tests remain extremely rare. In fact, independent tests of any kind of quantitative prediction are rare in ecology, spurring recent calls for a greater emphasis on prediction (Mouquet, Houlahan, Petchey). These authors argue that testing predictions will not only advance applied goals, such as forecasting ecological impacts of global change, but will also accelerate progress in basic research.    }

\new{ In the spirit of this argument, } we conducted a removal experiment to test whether our sagebrush steppe community model, which is based on observational data, underestimates the strength of interspecific competition and competitive release. We studied the responses of three perennial grass species to removal of sagebrush, and the response of sagebrush to removal of all grasses. The primary purpose of the experiment was not to directly measure competitive release, which is difficult to do over a short time period in a semiarid ecosystem with low and variable plant cover, but rather \new{to test whether our model accurately characterizes} the interspecific interactions that ultimately determine the magnitude of competitive release. We addressed two research questions: First, are the survival, growth or recruitment of our target species significantly different in control vs. removal treatments after accounting for differences in local species composition? Because our demographic models include the effects of neighborhood crowding, they should be able to predict how experimental removal of neighbors alters demographic performance. If the full effect of experimental removals is stronger than predicted, it means that our statistical models underestimate \new{the response to decreased crowding}. Our second question integrates across the full life cycle: how do short and long-term population-level responses to removals compare to projections from multispecies population models based on observational data? If models based on observational data adequately describe facilitative and competitive interactions, the population dynamics they project should be similar to projections from models that include information from the experimental removal treatments, a result which would validate the use of models and long-term data sets to study interspecific interactions.

@@

As in our previous studies of this system \citep{adler_coexistence_2010,adler_forecasting_2012}, the kernel is constructed from the fitted survival ($S$), growth ($G$), and recruitment ($R$) models, 
\begin{equation}
k_j(v,u,\vec{n})=S_j(u,\vec{n})G_j(v,u,\vec{n})+R_j(v,u,\vec{n}) 
\label{eqn:IPM} 
\end{equation}
(see Appendix \ref{AppendixA} for a complete description). The IPM uses a mean-field approximation for
neighborhood competition that captures the essential features of spatially-explicit local interactions; 
in Appendix \ref{sec:kernelMethods} we explain how the spline competition kernels were used 
to compute neighborhood competition as a function of individual size for the IPM.

@@

 \begin{figure}[tbp]
 \centering
 \includegraphics[width=1\textwidth]{cover_projections_1step}
 \caption{Observed and predicted cover for each of the four modeled species. Solid black lines show mean observed cover averaged across control (closed symbols) and removal treatment (open symbols) quadrats. Blue lines and symbols show one-step-ahead predictions from the baseline IBM (no removal treatment coefficients), averaged across quadrats. Red lines and symbols show one-step-ahead predictions from an IBM that includes removal treatment coefficients. \new{For \textit{H. comata}, lines showing predictions for the baseline and removal effect models are virtually indistinguishable.} Source file: \texttt{ibm\textbackslash summarize\textunderscore sims1step.r}. }
 \label{fig:IBM1step}
 \end{figure}

@@

% not sure yet where this should go
\new{Our study represents a very strong test of a surprising prediction for two reasons. First, the prediction was already published; the experiment 
represented a very real chance of getting it wrong...Second, we were just testing an observationally based model with new observations, but rather with
an experimental manipulation, increasing the potential to confront the model with a novel situation outside the historical range of variation... Add Steve's point about testing population models--in contrast to the Crone study, we do well when we control for environmental variation with random year effects...}

@@

Evidence for weak interspecific interactions and limited competitive release has interesting implications. If most 
interspecific interactions are weak, then single-species models may be adequate for forecasting population responses 
to environmental change \citep{adler_forecasting_2012,kleinhesselink_indirect_2015,Chu2016}. Similarly, differences among 
species in relative abundance should reflect variation in the strength of intraspecific density dependence more than 
variation in the strength of interspecific interactions \citep{yenni_strong_2012}. A recent global study showing little 
variation in the abundance of invasive grassland species in native and invaded ranges on different 
continents  \citep{firn_abundance_2011} is consistent with the idea that intraspecific interactions are the primary determinant 
of species abundances. In systems where this is true, effective management and conservation requires understanding the sources 
and species-specificity of intraspecific density dependence. A useful first step in this direction would be to test 
whether our model is also successful at predicting the effects of release from intraspecific competition.
